---
title: "phenotype_prelim"
author: "m quigg"
date: "2026-07-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Phenotype Preliminary Analysis
The purpose of this is to take a look at the resistance phenotype data the the USFS provided. They inoculated the trees with EAB eggs, then looked at the number of larvae that were remaining. 

##Goal 
See the distribution of the data to get an idea of what would be good covariables for the GWAS.

##Prep
```{r, include=FALSE}
###activate packages
library(tidyverse)
library(patchwork)
library(AICcmodavg)

###set working directory
setwd("C:/Users/mquig/Desktop/ash/GWAS_tails/phenotype")

###import the file
pheno_data=read.csv("LA-ProgenyTallyQCSAS_71426.csv", header = T)
```

After chatting with Zane, we realized that I was looking at all individuals, not just the ones we sequenced. So now, I want to rerun everything and incorporate it.
```{r}
###import the file
inds_seq=read.csv("USFS_Tails_Metadata_2023.xlsx - Sex_IDs_11.22.25.csv", header = T)

###redo the row names
inds_seq=inds_seq %>% 
  rename("IndShortName"=X.rname)

###join the dataframes
pheno_data=left_join(inds_seq, pheno_data, by = "IndShortName")

###remove the inds that we have sequence data for but no phenotype data
pheno_data=pheno_data %>% 
  drop_na(pL34)
```


##USFS Relevant Stats

###Proportion Host Kill (pHK or pTK)
This is the number of larvae that host killed divided by the number of good eggs.
```{r}
###basic stats
##max killed?
max(pheno_data$pHK)
#0.9166667
##min killed
min(pheno_data$pHK)
#0=none

##average
mean(pheno_data$pHK)
#0.2391576

###distribution
hist(pheno_data$pHK)
#right skewed

pHK_plot=ggplot(pheno_data, aes(x = pHK)) +
  geom_histogram(bins = 10, color = "magenta4", fill = "pink") +
  labs(title = "Proportion of Host Kills") +
  theme_classic()
pHK_plot
```
Ok so hist() broke it into bins with 0.1 intervals. There are a lot of individuals in the 0-0.1 pHK category.

###Realized Dose
This is the number of good eggs per tree.
```{r}
max(pheno_data$RlzDose)
min(pheno_data$RlzDose)
hist(pheno_data$RlzDose)
RlzDose_plot=ggplot(pheno_data, aes(x = RlzDose)) +
  geom_histogram(bins = 10, color = "aquamarine3", fill = "aquamarine") +
  labs(title = "Realized Dose of Good Eggs") +
  theme_classic()
RlzDose_plot
```

###Proportion 3 and 4 Instar Larvae
This is the proportion of recovered larvae that made it to instars 3 and 4. This is a proxy for EAB survival.
```{r}
max(pheno_data$pL34)
min(pheno_data$pL34)
hist(pheno_data$pL34)
pL34_plot=ggplot(pheno_data, aes(x = pL34)) +
  geom_histogram(bins = 10, color = "purple4", fill = "mediumpurple1") +
  labs(title = "Proportion of Larvae at Instars 3 or 4") +
  theme_classic()
pL34_plot
```

###Large Larvae
This is the number of large, surviving larvae at instars 3 and 4.
```{r}
max(pheno_data$LgLarv)
min(pheno_data$LgLarv)
hist(pheno_data$LgLarv)
LgLarv_plot=ggplot(pheno_data, aes(x = LgLarv)) +
  geom_histogram(bins = 10, color = "yellowgreen", fill = "darkolivegreen1") +
  labs(title = "Number of Large Larvae") +
  theme_classic()
LgLarv_plot
```
###Make a figure
```{r}
(pHK_plot + pL34_plot)/(RlzDose_plot + LgLarv_plot)
```
###Comparisons
Meg suggested that I run some little analyses to see if any of these variables are associated.

####pHK and pL34
```{r}
###pHK vs pL34
##make the graph
pHK_vs_pL34_plot=ggplot(pheno_data, aes(x = pHK, y = pL34)) +
  geom_point() +
  geom_smooth(method = "lm", color = "pink", fill = "magenta4") +
  theme_classic()
pHK_vs_pL34_plot

##make the model
pHK_vs_pL34_lm=lm(pL34 ~ pHK, pheno_data)
#intercept=0.8062
#-0.8184


###pL34 vs pHK
##make the graph
pL34_vs_pHK_plot=ggplot(pheno_data, aes(x = pL34, y = pHK)) +
  geom_point() +
  geom_smooth(method = "lm", color = "aquamarine3", fill = "aquamarine") +
  theme_classic()
pL34_vs_pHK_plot

##make the model
pL34_vs_pHK_lm=lm(pHK ~ pL34, pheno_data)
#intercept=0.7851
#-0.8748
```

###pHK and RlzDose
```{r}
###pHK vs RlzDose
##make the graph
pHK_vs_RlzDose_plot=ggplot(pheno_data, aes(x = pHK, y = RlzDose)) +
  geom_point() +
  geom_smooth(method = "lm", color = "pink", fill = "magenta4") +
  theme_classic()
pHK_vs_RlzDose_plot

##make the model
pHK_vs_RlzDose_lm=lm(RlzDose ~ pHK, pheno_data)
pHK_vs_RlzDose_lm
#intercept=388.11
#33.58

###RlzDose vs pHK
##make the graph
RlzDose_vs_pHK_plot=ggplot(pheno_data, aes(x = RlzDose, y = pHK)) +
  geom_point() +
  geom_smooth(method = "lm", color = "purple4", fill = "mediumpurple1") +
  theme_classic()
RlzDose_vs_pHK_plot

##make the model
Rlzdose_vs_pHK_lm=lm(pHK ~ RlzDose, pheno_data)
Rlzdose_vs_pHK_lm
#intercept=-0.137213
#0.001053
```

####RlzDose and pL34
```{r}
###pHK vs pL34
##make the graph
RlzDose_vs_pL34_plot=ggplot(pheno_data, aes(x = RlzDose, y = pL34)) +
  geom_point() +
  geom_smooth(method = "lm", color = "purple4", fill = "mediumpurple1") +
  theme_classic()
RlzDose_vs_pL34_plot

##make the model
RlzDose_vs_pL34_lm=lm(pL34 ~ RlzDose, pheno_data)
RlzDose_vs_pL34_lm
#intercept=1.000288
#-0.001067


###pL34 vs pHK
##make the graph
pL34_vs_RlzDose_plot=ggplot(pheno_data, aes(x = pL34, y = RlzDose)) +
  geom_point() +
  geom_smooth(method = "lm", color = "aquamarine3", fill = "aquamarine") +
  theme_classic()
pL34_vs_RlzDose_plot

##make the model
pL34_vs_RlzDose_lm=lm(RlzDose ~ pL34, pheno_data)
pL34_vs_RlzDose_lm
#intercept=418.52
#-36.39
```
###compare AIC scores
```{r}
AIC(pHK_vs_pL34_lm, pHK_vs_RlzDose_lm, pL34_vs_pHK_lm, pL34_vs_RlzDose_lm, Rlzdose_vs_pHK_lm, RlzDose_vs_pL34_lm)
```
I don't think this is really telling us anything.

###Make a figure
```{r}
(pHK_vs_pL34_plot + pHK_vs_RlzDose_plot) / (pL34_vs_pHK_plot + pL34_vs_RlzDose_plot) / (RlzDose_vs_pHK_plot + RlzDose_vs_pL34_plot)
```



##Other Relevant Visuals
I want to make a stacked bar chart with the good and bad eggs per tree. This is going to have a lot of bars, but I think it'll work.
```{r}
###create the data frame
egg=pheno_data %>% 
  dplyr::select(IndShortName, GoodEgg, BADEGG) %>% 
  rename("Good" = GoodEgg, "Bad" = BADEGG)
egg_long=pivot_longer(egg, cols = c(Good, Bad), names_to = "egg", values_to = "count")

###graph it
ggplot(egg_long, aes(x = IndShortName, y = count, fill = egg)) +
  geom_bar(position="stack", stat="identity")
```
So that doesn't look like what I wanted, but it works!


###Parents
```{r}
summary_families=pheno_data %>% 
  group_by(Family) %>% 
  summarise(n=length(Family)) %>% 
  ungroup()
summary_moms=pheno_data %>% 
  group_by(FemPar) %>% 
  summarise(n=length(FemPar)) %>% 
  ungroup()
summary_dads=pheno_data %>% 
  group_by(MalePar) %>% 
  summarise(n=length(MalePar)) %>% 
  ungroup()

family_plot=ggplot(summary_families, aes(x = Family, y = n)) +
  geom_bar(stat = "identity", color = "purple4", fill = "mediumpurple1") +
  labs(title = "Number per of Trees per Family", y = "count", x = "") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
family_plot

female_plot=ggplot(summary_moms, aes(x = FemPar, y = n)) +
  geom_bar(stat = "identity", color = "magenta4", fill = "pink") +
  labs(title = "Progeny from Each Female Parent", y = "count", x = "") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
female_plot

male_plot=ggplot(summary_dads, aes(x = MalePar, y = n)) +
  geom_bar(stat = "identity", color = "steelblue", fill = "lightskyblue1") +
  labs(title = "Progeny from Each Male Parent", y = "count", x = "") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
male_plot

###make a figure
family_plot/(female_plot+male_plot)
```

##Family Values <3
Now, I want to show how the phenotyping values vary by family.

###pHK
```{r}
family_pHK=pheno_data %>% 
  group_by(Family) %>% 
  dplyr::summarise(n=sum(!is.na(pHK)), 
                   mean=mean(pHK), 
                   sd=sd(pHK), 
                   se=sd/sqrt(n), 
                   lower.CI=mean-(1.96*se), 
                   upper.CI=mean+(1.96*se)) %>% 
  ungroup()

family_pHK_plot=ggplot(family_pHK, aes(x = Family, y = mean)) +
  geom_line(size = 1.5, color = "magenta4") +
  geom_errorbar(aes(ymin = lower.CI, ymax = upper.CI),
                width = 0.2, color = "magenta4") +
  geom_point(size = 3, shape = 21, fill = "pink", color = "magenta4") +
  labs(x = "Family", y = "pHK") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1), legend.position = "none")
family_pHK_plot

ggplot(pheno_data, aes(x = pHK)) +
  geom_histogram(color = "magenta4", fill = "pink", bins = 10) +
  facet_wrap(~ Family) +
  labs(y = "Count", x = "pHK") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

###pL34
```{r}
family_pL34=pheno_data %>% 
  group_by(Family) %>% 
  dplyr::summarise(n=sum(!is.na(pL34)), 
                   mean=mean(pL34), 
                   sd=sd(pL34), 
                   se=sd/sqrt(n), 
                   lower.CI=mean-(1.96*se), 
                   upper.CI=mean+(1.96*se)) %>% 
  ungroup()

family_pL43_plot=ggplot(family_pL34, aes(x = Family, y = mean)) +
  geom_line(size = 1.5, color = "purple4") +
  geom_errorbar(aes(ymin = lower.CI, ymax = upper.CI),
                width = 0.2, color = "purple4") +
  geom_point(size = 3, shape = 21, fill = "mediumpurple1", color = "purple4") +
  labs(x = "Family", y = "pL34") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1), legend.position = "none")
family_pL43_plot

ggplot(pheno_data, aes(x = pL34)) +
  geom_histogram(color = "purple4", fill = "mediumpurple1", bins = 10) +
  facet_wrap(~ Family) +
  labs(y = "Count", x = "pL34") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

###RlzDose
```{r}
family_RlzDose=pheno_data %>% 
  group_by(Family) %>% 
  dplyr::summarise(n=sum(!is.na(RlzDose)), 
                   mean=mean(RlzDose), 
                   sd=sd(RlzDose), 
                   se=sd/sqrt(n), 
                   lower.CI=mean-(1.96*se), 
                   upper.CI=mean+(1.96*se)) %>% 
  ungroup()

family_RlzDose_plot=ggplot(family_RlzDose, aes(x = Family, y = mean)) +
  geom_line(size = 1.5, color = "aquamarine3") +
  geom_errorbar(aes(ymin = lower.CI, ymax = upper.CI),
                width = 0.2, color = "aquamarine3") +
  geom_point(size = 3, shape = 21, fill = "aquamarine", color = "aquamarine3") +
  labs(x = "Family", y = "RlzDose") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1), legend.position = "none")
family_RlzDose_plot
```

###LgLarv
```{r}
family_LgLarv=pheno_data %>% 
  group_by(Family) %>% 
  dplyr::summarise(n=sum(!is.na(LgLarv)), 
                   mean=mean(LgLarv), 
                   sd=sd(LgLarv), 
                   se=sd/sqrt(n), 
                   lower.CI=mean-(1.96*se), 
                   upper.CI=mean+(1.96*se)) %>% 
  ungroup()

family_LgLarv_plot=ggplot(family_LgLarv, aes(x = Family, y = mean)) +
  geom_line(size = 1.5, color = "yellowgreen") +
  geom_errorbar(aes(ymin = lower.CI, ymax = upper.CI),
                width = 0.2, color = "yellowgreen") +
  geom_point(size = 3, shape = 21, fill = "darkolivegreen1", color = "yellowgreen") +
  labs(x = "Family", y = "LgLarv") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1), legend.position = "none")
family_LgLarv_plot
```

###figure
```{r}
family_pHK_plot / family_pL43_plot
```

##Species Distributions
Now, I want to take a look at the absolute disaster that is the species groupings.

###Add the species info
I created a column called "Species-ish" which lists if it is a hybrid and what kind. There are a lot of combos here.
```{r}
###import the data
species=read.csv("USFS_Tails_Metadata_2023.xlsx - Parentage_Species_IDs(Marne'sVersion).csv", header = T)

###add of the species info
species_pheno_data=left_join(pheno_data, species, by = "Family")
```

###pHK
```{r}
species_pHK=species_pheno_data %>% 
  group_by(Species.ish) %>% 
  dplyr::summarise(n=sum(!is.na(pHK)), 
                   mean=mean(pHK), 
                   sd=sd(pHK), 
                   se=sd/sqrt(n), 
                   lower.CI=mean-(1.96*se), 
                   upper.CI=mean+(1.96*se)) %>% 
  ungroup()

species_pHK_plot=ggplot(species_pHK, aes(x = Species.ish, y = mean)) +
  geom_line(size = 1.5, color = "magenta4") +
  geom_errorbar(aes(ymin = lower.CI, ymax = upper.CI),
                width = 0.2, color = "magenta4") +
  geom_point(size = 3, shape = 21, fill = "pink", color = "magenta4") +
  labs(x = "Species", y = "pHK") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1), legend.position = "none")
species_pHK_plot

species_pHK_facet=ggplot(species_pheno_data, aes(x = pHK)) +
  geom_histogram(color = "magenta4", fill = "pink", bins = 10) +
  facet_wrap(~ Species.ish) +
  labs(y = "Count", x = "pHK") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
species_pHK_facet
```

###pL34
```{r}
species_pL34=species_pheno_data %>% 
  group_by(Species.ish) %>% 
  dplyr::summarise(n=sum(!is.na(pL34)), 
                   mean=mean(pL34), 
                   sd=sd(pL34), 
                   se=sd/sqrt(n), 
                   lower.CI=mean-(1.96*se), 
                   upper.CI=mean+(1.96*se)) %>% 
  ungroup()

species_pL43_plot=ggplot(species_pL34, aes(x = Species.ish, y = mean)) +
  geom_line(size = 1.5, color = "purple4") +
  geom_errorbar(aes(ymin = lower.CI, ymax = upper.CI),
                width = 0.2, color = "purple4") +
  geom_point(size = 3, shape = 21, fill = "mediumpurple1", color = "purple4") +
  labs(x = "Species", y = "pL34") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1), legend.position = "none")
species_pL43_plot

species_pL34_facet=ggplot(species_pheno_data, aes(x = pL34)) +
  geom_histogram(color = "purple4", fill = "mediumpurple1", bins = 10) +
  facet_wrap(~ Species.ish) +
  labs(y = "Count", x = "pL34") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
species_pL34_facet
```

###making figures
```{r}
(species_pHK_plot + species_pL43_plot) / (species_pHK_facet + species_pL34_facet)
```
That's not a great figure but they look interesting on their own.

##What do we need?
So we have a fair number of individuals that don't have phenotype or species data. I need to identify what individuals were missing phenotype data for, and which families don't have species data.

###No phenotype data
```{r}
###join the dataframes
pheno_data_withNA=left_join(inds_seq, pheno_data, by = "IndShortName")

###remove the inds that we have sequence data for but no phenotype data
missing_pheno=pheno_data_withNA %>% 
  filter(is.na(pHK)) %>% 
  dplyr::select(IndShortName)

###save the csv
write.csv(missing_pheno, "C:/Users/mquig/Desktop/ash/GWAS_tails/phenotype/missing_phenotypes.csv")
```

###no species data
```{r}
missing_species=species_pheno_data %>% 
  filter(is.na(Species.ish)) %>% 
  dplyr::select(IndShortName, Family)
```
